Initial estuarine response to the nutrient-rich Piney Point release into Tampa Bay, Florida

### 2022-01-30 revision/review_response.Rmd

We sincerely thank the reviewers for providing comments on our manuscript.  We provide a point-by-point response to these comments below. In general, most of the revisions address concerns raised by all three reviewers, including 1) improvement in writing style for clarity and brevity, and 2) additional information on sampling methods.  We have also addressed concerns of Reviewer 2 about potential nutrient cycling by adding a section in the discussion and concerns of Reviewer 4 regarding additional detail on the simulation modelling.

## Reviewer 1

The authors (a very long list of authors) have described the sequence of events that followed a phosphate mining spill. The paper is long and tedious- and in the end cannot conclude any strong relationships with red tide occurrence. I wish I could be more positive about this paper, but it has the feeling that authors were rushing to get something in press, rather than taking the time to write a well crafted manuscript. There are some very good writers on the list of authors- so I am a bit surprised that they all agreed to this final version.

* __Response__: Thank you for your comments. We have made revisions to improve the writing, by reducing length in many locations and making additional edits where noted. Note that many of the edits in response to the more specific comments from the other reviewers are used to address some of the comments below.

@@

stats section- poorly written, confusing, wordy

* __Response__: Please see the responses to reviewer 2.

Results- rambling, disorganized and non-quantitative. Paragraphs don't make sense (try using topic sentences and structuring concepts accordingly)

* __Response__: Please see the responses to reviewer 2, particularly regarding the first section of the results.

@@

Discussion- I was left wondering what I learned. No conclusions drawn

* __Response__: Please see the responses to reviewer 2, particularly regarding the edits to the discussion sections.

@@

In all, it seemed like an interesting exercise to describe all the patterns, but this does not make a quality manuscript that will stand the test of time.

* __Response__: We are confident that the edits made in response to your comments and those from the other reviewers have made this a more compelling and valuable contribution.

### 2022-01-31 manu-draft.Rmd
revisions done, need to proof and format for resubmission

From March 30th to April 9th, 2021, 814 million liters of legacy phosphate mining wastewater and marine dredge water from the Piney Point facility were released into lower Tampa Bay (Florida, USA). This resulted in an estimated addition of 186 metric tons of total nitrogen, exceeding typical annual external nitrogen load estimates to lower Tampa Bay in a matter of days. An initial phytoplankton bloom (non-harmful diatoms) was first observed in April in the lower Bay. Filamentous cyanobacteria blooms (*Dapis* spp.) peaked in June, followed by a bloom of the red tide organism *Karenia brevis*. Reported fish kills tracked *K. brevis* concentrations, prompting cleanup of over 1600 metric tons of dead fish. Seagrasses had minimal changes over the study period.  By comparing these results to baseline environmental monitoring data, we conclude that many of the biological responses observed after the release from Piney Point are abnormal relative to historic conditions.

@@

Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources [@Tayibi09;@Hudson11]. Phosphate fertilizer is produced through the "wet process" reaction to create phosphoric acid by treating mined phosphate rock with sulfuric acid [@Burnett01;@Perez16].  The process generates large amounts of waste, creating approximately one unit of phosphoric acid per five units of phosphogypsum waste (CaSO$_4\cdot$ H$_2$O).  This waste is typically stored on-site in large earthen stacks (gypstacks) capable of holding millions of liters of process water.  Water quality in gypstacks can vary depending on processing method used at the mining facility, background geological characteristics of the region, and on-site practices for managing stormwater or other activities that can introduce additional materials to the holding ponds [@Henderson04;@Perez10].  In addition to elevated phosphorus concentrations, other nutrients, contaminants, and radionuclides may be present at values much higher than those of natural surface waters [@Burnett01;@Beck18b]. Many of these gypstacks no longer support active mining and aging infrastructure combined with climate change and seasonal stressors (e.g., heavy precipitation events) have reduced the capacity of these facilities to maintain water on site. Environmental and human health risks associated with these stacks can occur through controlled or uncontrolled releases to surface waters or groundwater contamination through leaching from unlined or poorly maintained stacks [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].

Ongoing threats and challenges to protecting water quality of northern Gulf of Mexico estuaries persist despite recent environmental recovery.  Successful restoration has been accomplished through ecosystem management paradigms that are based primarily on the control of nutrient pollutants from atmospheric deposition, stormwater, and wastewater.  Nitrogen inputs from external sources are well understood as drivers of algal blooms in coastal environments that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14]. Seagrasses in particular are valuable habitat-defining species that provide many ecosystem services, but are especially vulnerable to the impacts of nutrient pollution on water quality because of cascading negative effects of nitrogen, phytoplankton growth and persistence, water clarity, light limitation, and epiphyte loading on seagrass growth and survival [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].  Historical gains in seagrass coverage in southwest Florida estuaries have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads delivered to surface waters [@Janicki96;@Greening16;@Tomasko18;@Tomasko20].  Together, these efforts have resulted in the long-term recovery of Tampa Bay and other Gulf Coast estuaries through a reduction in external nitrogen loads, improvements in water clarity, and baywide expansion of seagrass coverage to benchmark targets established for the region [@Greening14;@Sherwood17]. The three contiguous National Estuary Programs of southwest Florida (Tampa Bay, Sarasota Bay, Charlotte Harbor) and their partners have been instrumental in coordinating efforts to address legacy pollutants and current threats to the long-term protection of estuarine resources.

The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer to support agricultural production [@Henderson04].  By 2001, an estimated 36 million metric tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].  Currently, seventeen phosphogypsum stacks exist in the Tampa Bay watershed.  Effective management and final closure of these facilities are imperative to reduce threats to prior ecosystem recovery efforts and investments. The Piney Point facility located in Palmetto, Florida is a large, remnant gypstack with three holding ponds located 3 kilometers from the shore of Tampa Bay and near two Florida Aquatic Preserves [see supplement for a history of the facility; @Henderson04].  Holding capacity of the ponds has decreased over time from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee.  Releases from the stacks occurred in the early 2000s and in 2011 to nearby Bishop Harbor connected to Tampa Bay.  Those releases resulted in spatially-restricted, ecosystem responses including localized harmful algal blooms and increased macroalgal abundance [@Switzer11;@Garrett11].

More recently, leakages were detected from a tear in the plastic liner of the southern holding pond (NGS-S) at Piney Point.  In response, the Florida Department of Environmental Protection (FDEP) authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 29th, 2021 to release water from the southern gypstack directly into lower Tampa Bay to prevent catastrophic failure of the facility. At that time, approximately 1.8 billion liters of mixed legacy phosphate mining wastewater and seawater from port dredging operations were being held in the failing gypstack. Piney Point historically produced Diammonium Phosphate ((NH$_4$)$_2$HPO$_4$) and the remnant stackwater has very high concentrations of both phosphorus and nitrogen.  Water quality parameters of NGS-S measured in 2019 showed total phosphorus (160 mg/L) and total nitrogen (230 mg/L) were approximately three orders of magnitude higher than typical concentrations in Tampa Bay. From March 30th to April 9th, approximately 814 million liters (215 million gallons) of stack water were released under FDEP authorization to lower Tampa Bay.  Over this ten day period, an estimated 186 metric tons (205 tons) of nitrogen were delivered to the bay, exceeding contemporary annual estimates of external nutrient loads to lower Tampa Bay in a matter of days [@tbep0417].

This paper provides an initial assessment of environmental conditions in Tampa Bay over the six month period after the release of legacy phosphate mining wastewater from the Piney Point facility in 2021.  The goal is to describe the results of monitoring data of surface waters collected in response to the event to assess relative deviation of current conditions from long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region. Numerous studies, as well as the successful nutrient management paradigm for Tampa Bay, have demonstrated nitrogen-limitation in Tampa Bay and the system is generally considered phosphorus enriched [@Wang99;@Greening06;@Greening14]. As such, we focus on nitrogen in our analyses as the identified limiting nutrient for Tampa Bay and its potential to create water quality conditions unfavorable for seagrass growth due to enhanced algal production.  A timeline of events is provided, which is supported by the quantitative results from 2021 routine and response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency releases from the Piney Point facility.  The results from this paper provide examples of anticipated short-term responses to acute nutrient loadings from legacy mining facilities in the broader context of historical conditions that may influence system response to these events.

@@

Differences in observations between months for April to September for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13]. These tests were used to statistically characterize the temporal progression of changes in the bay following release from Piney Point.  Probability values were adjusted using the sequential Bonferroni method described in [@Holm79] to account for the increased probability of Type I error rates with multiple comparisons. An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.  For water quality variables, monthly averages from long-term monitoring data were subtracted from 2021 observations to account for normal seasonal variation not attributed to potential effects from Piney Point.  Similar corrections were not done for monthly comparisons of seagrass and macroalgae data because comparable long-term seasonal data do not exist. Frequency occurrence estimates were used to evaluate macroalgae and seagrasses as a standard metric used in previous analyses in Tampa Bay [@Johansson16;@Sherwood17]. Methods used to accommodate measured concentrations of water quality variables that were below detection included summary statistics (e.g., median, mean, and standard deviation) following estimates of the empirical cumulative distribution functions for each parameter using the Kaplan-Meier method for censored data [@Helsel05;@Lee20].

The R statistical programming language (v4.0.2) was used for all analyses [@RCT20].  We imported data using the googlesheets4 [@Bryan20] and googledrive [@DAgostino20] R packages and used tidyverse [@Wickham19] packages to format data for analysis.  The tbeptools R package [@Beck21] was used to import and summarize long-term monitoring data (EPC water quality data and seagrass transect data). The NADA R package [@Lee20] was used for analysis of censored data. All spatial analyses were done using the simple features (sf) R package [@Pebesma18]. The mgcv R package [@Wood17] was used to create the GAMs for water quality parameters. All datasets used in this study are available from an open access data archive hosted on the Knowledge Network for Biocomplexity [@Beck21d].  Materials for reproducing the analyses, figures, tables, and other content in this paper are provided in a GitHub repository. Finally, the Piney Point Environmental Monitoring Dashboard can be used to view all data included in this paper through an interactive, online application [@Beck21c].  Links and details are provided in supplement.

@@

Water quality conditions in the northern gypstack measured in 2019 and measured directly at the point of discharge in 2021 showed concentrations that were generally much higher for key water quality parameters as compared to baseline conditions in Tampa Bay (Table \@ref(tab:stktab)).  Notably, total ammonia nitrogen was measured at 210 mg/L at Piney Point and in the discharge, compared to a long-term median of 0.02 mg/L in lower Tampa Bay. Similar differences for total phosphorus, TN, and chl-a were observed when comparing stack conditions with those of the ambient conditions in Tampa Bay.  These contrasts provided a context for the monitoring data collected in 2021.

Samples collected in the bay between April through September 2021 indicated that water quality conditions were unusual and outside of normal values expected for each month. A total of `r nwq` samples were collected and analyzed for chl-a, dissolved oxygen, TN, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, Secchi depth, and temperature (Table \@ref(tab:wqsumtab)). The percentage of observations outside of the normal range (mean +/- 1 standard deviation from long-term data) varied by location and parameter.  For chl-a, `r chlabv[['1']]`\% of the observations from April through September were above the normal range for Area 1 located closest to the discharge point, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for Areas 2 (to the north) and 3 (to the south), respectively. Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in Area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in Area 2 and `r tnabv[['3']]`\% in Area 3.  Secchi observations were below the normal range for `r secblw[['1']]`\% of observations in Area 1 and for `r secblw[['2']]`\% and `r secblw[['3']]`\% of observations in Areas 2 and 3.  Notable differences were also observed for dissolved oxygen (e.g., `r dosabv[['1']]`\% were above in Area 1, `r dosabv[['2']]`\% in Area 2).  Physical parameters (salinity, temperature) and inorganic nitrogen (ammonia, nitrate/nitrite) were more often in normal ranges, although initial time series showed much higher concentrations for ammonia in April near Area 1.  Ammonia concentrations near the point of discharge were observed in excess of 10 mg/L, about three orders of magnitude above baseline (Figures S2, S3), similar to the discharge measurements in Table \@ref(tab:stktab)). Inorganic nitrogen did not persist at high concentrations past April as it was likely rapidly utilized by phytoplankton (see below).  Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi depth) for many locations near Piney Point (Area 1), Anna Maria Sound (Area 3), and the northern mouth of Tampa Bay (Area 3, Figure \@ref(fig:wqmap)).

Total nitrogen, chl-a, and Secchi depth followed temporal progressions in 2021 that were distinct from long-term seasonal trends estimated from historical data (Figure \@ref(fig:wqgam)).  For Area 1, TN and chl-a concentrations were frequently above normal ranges during April. Chlorophyll-a concentrations were observed in excess of 50 $\mu$g/L, altough median concentrations for each week in April were less than 10 $\mu$g/L.  The initial chlorophyll peak was associated with a localized phytoplankton bloom generally dominated by diatoms. The initial diatom bloom did not persist past April. Concentrations decreased slightly until June and July when values increased again above the seasonal expectation, coincident with the increase in *K. brevis* concentrations. Many Secchi observations in Area 1 were lower than normal in April and July.  Observations in Areas 2 and 3 were more often within the normal seasonal range, with some exceptions for TN and chl-a in Area 3 in  April, May, and July.  Statistical comparisons between months for seasonally-corrected observations of TN, chl-a, and Secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqgam). Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant (*p* < 0.05) for TN, chl-a, and Secchi depth for Areas 1 and 3 and for TN and chl-a for Area 2 (Table \@ref(tab:wqcmptab)). Further analysis with multiple comparison tests generally showed that April/May were different from June/July depending on Area and parameter, such that observations in the later months were generally higher (or lower for Secchi) corresponding to increasing *K. brevis* abundances by mid-summer.

@@

A typical pattern for macroalgae and seagrass observed at many of the transects is shown in Figure \@ref(fig:trnex). Transect S3T6 is located less than one kilometer to the north of Port Manatee. Macroalgal abundances changed over the course of sampling similar to the remainder of transects sampled during the study.  Red macroalgae were present in high abundances from April to May.  Filamentous cyanobacteria (*Dapis* spp.) mats were first observed on May 24th and was present at all of the sample locations along this transect on June 4th and 15th.  Filamentous cyanobacteria persisted through June and July, but was not observed in abundance after July 20th.  Green macroalgae taxa were first observed in July, although at generally low abundances. Red macroalgae were the dominant taxa by the end of September. Overall abundance of seagrass did not change from April 22nd through September. The site is dominated by manatee grass that was observed at nearly all of the sample points along the transect at varying coverages.

Monthly summaries in frequency occurrence by area (Figure \@ref(fig:trnfrq)) provided an indication of macroalgae and seagrass trends in 2021 across all transects.  No transects were sampled in Area 2 to the north of Piney Point and no transects were sampled past September in Area 1 given allocated sampling effort following projected dispersal patterns of the discharge from the TBCOM simulations.  Macroalgal dominance varied across the months, similar to the example in Figure \@ref(fig:trnex).  Red macroalgae was the dominant group across all months and areas, with the highest frequency occurrences observed in April (`r fomoarea1apr['Red']`\% in Area 1, `r fomoarea3apr['Red']`\% in Area 3). Reductions in red macroalgae frequency occurrence were observed in June when cyanobacteria frequency occurrence peaked, with greater coverage of cyanobacteria in Area 3 (`r fomoarea3jun['Cyanobacteria']`\%) compared to Area 1 (`r fomoarea1jun['Cyanobacteria']`\%).  Notable blooms of the filamentous cyanobacteria (*Dapis* spp.) were observed in Anna Maria Sound (Area 3) and near Port Manatee (Area 1) (Figure \@ref(fig:map)), typically observed covering benthic and seagrass habitats, in addition to large floating mats on the surface.  Green macroalgae had the second lowest frequency occurrence, although it increased slightly by the end of the study period (`r fomoarea1sep['Green']`\% in September in Area 1, `r fomoarea3oct['Green']`\% in October in Area 3).  Brown macroalgae was only observed at one transect. For seagrass, both areas had generally stable total frequency occurrence. Turtle grass (*T. testudinum*) occurred in higher frequency occurrence in both areas (`r fomoarea1ave['Thalassia testudinum']`\% overall in Area 1, `r fomoarea3ave['Thalassia testudinum']`\% overall in Area 3), compared to shoal grass (*H. wrightii*, `r fomoarea1ave['Halodule wrightii']`\% Area 1, `r fomoarea3ave['Halodule wrightii']`\% Area 3) and manatee grass (*S. filiforme*, `r fomoarea1ave['Syringodium filiforme']`\% Area 1, `r fomoarea3ave['Syringodium filiforme']`\% Area 3).  Slight changes in frequency occurrence in Area 3 were observed for all species starting in July, with a slight reduction in frequency occurrence of turtle grass and an increase in shoal grass and manatee grass. Statistical analyses with multiple comparison tests confirmed the general trends described above, with significant changes observed only for macroalgae (Tables S1, S2).  Tests using Braun Blanquet cover estimates confirmed the results from the frequency occurrence estimates (Tables S3, S4).

@@

In the broader context of mining impacts to surface waters, these results reinforce the understanding that legacy pollutants from phosphate mining can negatively affect environmental resources.  In addition to Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have been affected by pollutants from unanticipated gypstack releases.  Grand Bay is a 7500 hectare protected area in southern Mississippi that has been exposed to phosphorus-rich and highly acidic water from a defunct gypstack [@Dillon15;@Beck18b].  Two spills have occurred in Grand Bay, the first in 2005 following failure of the retaining walls after a heavy rain event and the second in 2012 after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.  Massive fish kills were observed and likely related to low pH of the water released.  Unlike Piney Point, inorganic nitrogen concentrations of the release were low due to a different fertilizer production method and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].  Phosphate loads to Tampa Bay from Piney Point were similar in magnitude to the nitrogen loads, although concentrations were within normal baseline ranges within a month after the release stopped (Figures S2, S3).  The fate of the phosphorus in Tampa Bay is less understood than that of nitrogen.  Regardless, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].  Legacy wastewater from fertilizer production has been poorly maintained at some facilities and long-term plans are insufficient to safely dispose of remnant pollutants that pose a risk of significant impacts to coastal resources that increases over time.  These are not isolated examples and enhanced regulatory oversight is needed to safely and effectively close these types of facilities [@Nelson21].

>>>>

Legacy mining facilities pose significant risks to aquatic resources.  From March 30th to April 9th, 2021, 814 million liters of phosphate mining wastewater and marine dredge water from the Piney Point facility were released into lower Tampa Bay (Florida, USA). This resulted in an estimated addition of 186 metric tons of total nitrogen, exceeding typical annual external nitrogen load estimates to lower Tampa Bay in a matter of days. An initial phytoplankton bloom (non-harmful diatoms) was first observed in April. Filamentous cyanobacteria blooms (*Dapis* spp.) peaked in June, followed by a bloom of the red tide organism *Karenia brevis*. Reported fish kills tracked *K. brevis* concentrations, prompting cleanup of over 1600 metric tons of dead fish. Seagrasses had minimal changes over the study period.  By comparing these results to baseline environmental monitoring data, we demonstrate adverse water quality changes in response to external nitrogen loads into Tampa Bay.

@@

Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources [@Tayibi09;@Hudson11]. 
The production of phosphate fertilizer generates large amounts of phosphogpysum waste (CaSO$_4\cdot$ H$_2$O) that is typically stored on-site in large earthen stacks (gypstacks) capable of holding millions of liters of process water.  Water quality in gypstacks can vary depending on processing method used at the mining facility, background geological characteristics of the region, and on-site practices for managing stormwater or other activities that can introduce additional materials to the holding ponds [@Henderson04;@Perez10].  In addition to elevated phosphorus concentrations, other nutrients, contaminants, and radionuclides may be present at values much higher than natural surface waters [@Burnett01;@Beck18b]. Many of these gypstacks no longer support active mining and aging infrastructure combined with climate change and seasonal stressors (e.g., heavy precipitation events) have reduced the capacity of these facilities to maintain water on site. Numerous studies have documented the environmental and human health risks associated with these stacks [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].

The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer to support agricultural production [@Henderson04].  By 2001, an estimated 36 million metric tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].  Effective management and final closure of these facilities are imperative to reduce threats to prior ecosystem recovery efforts and investments. The Piney Point facility located in Palmetto, Florida is a large, remnant gypstack with three holding ponds located 3 kilometers from the shore of Tampa Bay and near two Florida Aquatic Preserves [see supplement for a history of the facility; @Henderson04].  Holding capacity of the ponds has decreased over time from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee.  Releases from the stacks occurred in the early 2000s and in 2011 to nearby Bishop Harbor connected to Tampa Bay.  Those releases resulted in spatially-restricted, ecosystem responses including localized harmful algal blooms and increased macroalgal abundance [@Switzer11;@Garrett11].

In March 2021, leakages were detected from a tear in the plastic liner of the southern holding pond (NGS-S) at Piney Point.  At that time, approximately 1.8 billion liters of mixed legacy phosphate mining wastewater and seawater from port dredging operations were being held in the failing gypstack. Piney Point historically produced Diammonium Phosphate ((NH$_4$)$_2$HPO$_4$) and the remnant stackwater has very high concentrations of nitrogen, in addition to phosphorus.  Water quality parameters of NGS-S measured in 2019 showed total phosphorus (160 mg/L) and total nitrogen (230 mg/L) were approximately three orders of magnitude higher than typical concentrations in Tampa Bay. From March 30th to April 9th, approximately 814 million liters (215 million gallons) of stack water were released to lower Tampa Bay following an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) authorized by the Florida Department of Environmental Protection (FDEP).  Over this ten day period, an estimated 186 metric tons (205 tons) of nitrogen were delivered to the bay, exceeding contemporary annual estimates of external nutrient loads to lower Tampa Bay in a matter of days [@tbep0417].

This paper provides an initial assessment of environmental conditions in Tampa Bay over the six month period after the release of legacy phosphate mining wastewater from the Piney Point facility in 2021.  The goal is to describe the results of monitoring data of surface waters collected in response to the event to assess relative deviation of current conditions from long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region. Numerous studies, as well as the successful nutrient management paradigm for Tampa Bay, have demonstrated nitrogen-limitation in Tampa Bay and the system is generally considered phosphorus enriched [@Wang99;@Greening06;@Greening14].  As such, we focus on nitrogen in our analyses as the identified limiting nutrient for Tampa Bay and its potential to create water quality conditions unfavorable for seagrass growth due to enhanced algal production.  Our analysis evaluated datasets that are descriptive of the vulnerability of seagrasses to nutrient pollution though cascading negative effects of nitrogen, phytoplankton growth and persistence, and water clarity on seagrass growth and survival [@Dixon95;@Kenworthy96;@Greening06;@Beck18a]. A timeline of events is provided, which is supported by the quantitative results from 2021 routine and response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency releases from the Piney Point facility.  The results from this paper provide an unprecedented chronology of short-term estuarine response to acute nutrient loadings from legacy mining facilities, where context would not have been possible without the long-term monitoring datasets available for the region.

@@

Differences in observations between months for April to September for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13]. These tests were used to statistically characterize the temporal progression of changes in the bay following release from Piney Point, e.g., were July conditions significantly different from April.  Probability values were adjusted using the sequential Bonferroni method described in [@Holm79] to account for the increased probability of Type I error rates with multiple comparisons. An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.  For water quality variables, monthly averages from long-term monitoring data were subtracted from 2021 observations to account for normal seasonal variation not attributed to potential effects from Piney Point.  Similar corrections were not done for monthly comparisons of seagrass and macroalgae data because comparable long-term seasonal data do not exist. Frequency occurrence estimates were used to evaluate macroalgae and seagrasses as a standard metric used in previous analyses in Tampa Bay [@Johansson16;@Sherwood17]. Methods used to accommodate measured concentrations of water quality variables that were below detection included summary statistics (e.g., median, mean, and standard deviation) following estimates of the empirical cumulative distribution functions for each parameter using the Kaplan-Meier method for censored data [@Helsel05;@Lee20].

The R statistical programming language (v4.0.2) was used for all analyses [@RCT20].  We imported data using the googlesheets4 [@Bryan20] and googledrive [@DAgostino20] R packages and used tidyverse [@Wickham19] packages to format data for analysis.  The tbeptools R package [@Beck21] was used to import and summarize long-term monitoring data (EPC water quality data and seagrass transect data). The NADA R package [@Lee20] was used for analysis of censored data. All spatial analyses were done using the simple features (sf) R package [@Pebesma18]. The mgcv R package [@Wood17] was used to create the GAMs for water quality parameters. All datasets used in this study are available from an open access data archive hosted on the Knowledge Network for Biocomplexity [@Beck21d].  Materials for reproducing the analyses, figures, tables, and other content in this paper are provided in a [GitHub repository](https://github.com/tbep-tech/piney-point-manu). Finally, the Piney Point Environmental Monitoring Dashboard can be used to view all data included in this paper through an interactive, online application [@Beck21c].  Links and details are provided in supplement.

@@

Water quality conditions in the northern gypstack measured in 2019 and measured directly at the point of discharge in 2021 showed concentrations that were generally much higher for key water quality parameters as compared to baseline conditions in Tampa Bay (Table \@ref(tab:stktab)).  Notably, total ammonia nitrogen was measured at 210 mg/L at Piney Point and in the discharge, compared to a long-term median of 0.02 mg/L in lower Tampa Bay. Similar differences for total phosphorus, TN, and chl-a were observed when comparing stack conditions with those of the ambient conditions in Tampa Bay.

Samples collected in the bay between April through September 2021 indicated that water quality conditions were outside of normal values expected for each month. A total of `r nwq` samples were collected and analyzed for chl-a, dissolved oxygen, TN, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, Secchi depth, and temperature (Table \@ref(tab:wqsumtab)). The percentage of observations outside of the normal range (mean +/- 1 standard deviation from long-term data) varied by location and parameter.  For chl-a, `r chlabv[['1']]`\% of the observations from April through September were above the normal range for Area 1 located closest to the discharge point, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for Areas 2 (to the north) and 3 (to the south), respectively. Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in Area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in Area 2 and `r tnabv[['3']]`\% in Area 3.  Secchi observations were below the normal range for `r secblw[['1']]`\% of observations in Area 1 and for `r secblw[['2']]`\% and `r secblw[['3']]`\% of observations in Areas 2 and 3.  Notable differences were also observed for dissolved oxygen (e.g., `r dosabv[['1']]`\% were above in Area 1, `r dosabv[['2']]`\% in Area 2).  Physical parameters (salinity, temperature) and inorganic nitrogen (ammonia, nitrate/nitrite) were more often in normal ranges, although initial time series showed much higher concentrations for ammonia in April near Area 1.  Ammonia concentrations near the point of discharge were observed in excess of 10 mg/L in April, about three orders of magnitude above baseline (Figures S2, S3), similar to the discharge measurements in Table \@ref(tab:stktab). Inorganic nitrogen did not persist at high concentrations past April as it was likely rapidly utilized by phytoplankton (see below).  Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi depth) for many locations near Piney Point (Area 1), Anna Maria Sound (Area 3), and the northern mouth of Tampa Bay (Area 3, Figure \@ref(fig:wqmap)).

Total nitrogen, chl-a, and Secchi depth followed temporal progressions in 2021 that were distinct from long-term seasonal trends estimated from historical data (Figure \@ref(fig:wqgam)).  For Area 1, TN and chl-a concentrations were frequently above normal ranges during April. Chlorophyll-a concentrations were observed in excess of 50 $\mu$g/L, altough median concentrations for each week in April were less than 10 $\mu$g/L.  The initial chlorophyll peak was associated with a localized phytoplankton bloom generally dominated by diatoms. The initial diatom bloom did not persist past April. Chlorophyll concentrations decreased slightly until June and July when values increased again above the seasonal expectation, coincident with an increase in *K. brevis* concentrations to bloom levels. Many Secchi observations in Area 1 were lower than normal in April and July.  Observations in Areas 2 and 3 were more often within the normal seasonal range, with some exceptions for TN and chl-a in Area 3 in  April, May, and July.

Statistical comparisons between months for seasonally-corrected observations of TN, chl-a, and Secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqgam). Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant (*p* < 0.05) for TN, chl-a, and Secchi depth for Areas 1 and 3 and for TN and chl-a for Area 2 (Table \@ref(tab:wqcmptab)). Further analysis with multiple comparison tests generally showed that April/May were different from June/July depending on Area and parameter, such that observations in the later months were generally higher (or lower for Secchi) corresponding to increasing *K. brevis* abundances by mid-summer.

@@

A typical temporal pattern for macroalgae and seagrass observed at many of the transects is shown in Figure \@ref(fig:trnex), using transect S3T6 near Port Manatee as an example. Macroalgal abundances changed over the course of sampling similar to the remainder of transects sampled during the study.  Red macroalgae were present in high abundances from April to May.  Filamentous cyanobacteria (*Dapis* spp.) mats were first observed on May 24th and was present at all of the sample locations along this transect on June 4th and 15th.  Filamentous cyanobacteria persisted through June and July, but was not observed in abundance after July 20th.  Green macroalgae taxa were first observed in July, although at generally low abundances. Red macroalgae were the dominant taxa by the end of September. Overall abundance of seagrass did not change from April 22nd through September. The site is dominated by manatee grass that was observed at nearly all of the sample points along the transect at varying coverages.

Monthly summaries in frequency occurrence by area (Figure \@ref(fig:trnfrq)) provided an indication of macroalgae and seagrass trends in 2021 across all transects.  No transects were sampled in Area 2 to the north of Piney Point and no transects were sampled past September in Area 1 given allocated sampling effort following projected dispersal patterns of the discharge from the TBCOM simulations. Red macroalgae was the dominant group across all months and areas, with the highest frequency occurrences observed in April (`r fomoarea1apr['Red']`\% in Area 1, `r fomoarea3apr['Red']`\% in Area 3). Reductions in red macroalgae frequency occurrence were observed in June when cyanobacteria frequency occurrence peaked, with greater coverage of cyanobacteria in Area 3 (`r fomoarea3jun['Cyanobacteria']`\%) compared to Area 1 (`r fomoarea1jun['Cyanobacteria']`\%).  Notable blooms of the filamentous cyanobacteria (*Dapis* spp.) were observed in Anna Maria Sound (Area 3) and near Port Manatee (Area 1) (Figure \@ref(fig:map)), typically observed covering benthic and seagrass habitats, in addition to large floating mats on the surface.  Green macroalgae had the second lowest frequency occurrence, although it increased slightly by the end of the study period (`r fomoarea1sep['Green']`\% in September in Area 1, `r fomoarea3oct['Green']`\% in October in Area 3).  For seagrass, both areas had generally stable total frequency occurrence. Turtle grass (*T. testudinum*) occurred in higher frequency occurrence in both areas (`r fomoarea1ave['Thalassia testudinum']`\% overall in Area 1, `r fomoarea3ave['Thalassia testudinum']`\% overall in Area 3), compared to shoal grass (*H. wrightii*, `r fomoarea1ave['Halodule wrightii']`\% Area 1, `r fomoarea3ave['Halodule wrightii']`\% Area 3) and manatee grass (*S. filiforme*, `r fomoarea1ave['Syringodium filiforme']`\% Area 1, `r fomoarea3ave['Syringodium filiforme']`\% Area 3).  Slight changes in frequency occurrence in Area 3 were observed for all species starting in July, with a slight reduction in frequency occurrence of turtle grass and an increase in shoal grass and manatee grass. Statistical analyses with multiple comparison tests confirmed the general trends described above, with significant changes observed over time only for macroalgae (Tables S1, S2).  Tests using Braun Blanquet cover estimates confirmed the results from the frequency occurrence estimates (Tables S3, S4).

@@

In the broader context of mining impacts to surface waters, these results reinforce the understanding that legacy pollutants from phosphate mining can negatively affect environmental resources.  In addition to Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have been affected by pollutants from unanticipated gypstack releases.  For example, two spills have occurred in Grand Bay, Mississippi, the first in 2005 following failure of the retaining walls after a heavy rain event and the second in 2012 after passage of Hurricane Isaac when the holding capacity of the local gypstack was exceeded again with heavy rainfall [@Dillon15;@Beck18b].  The historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].  Legacy wastewater from fertilizer production has been poorly maintained at some facilities and long-term plans are insufficient to safely dispose of remnant pollutants that pose a risk of significant impacts to coastal resources that increases over time.  These are not isolated examples and enhanced regulatory oversight is needed to safely and effectively close these types of facilities [@Nelson21].  Local, regional, and state partners should continue to pursue management and policy actions that can mitigate the continued threats of these facilities to the health of coastal resources.  These efforts are critical to managing Gulf of Mexico ecosystems given past successes and the need to address ongoing threats of climate change, human population growth, habitat loss, severe weather events, and recurring pollutant sources.
